---
title: "Rmarkdown: Climate change drives loss of bacterial gut mutualists at expense of host survival in wild meerkats"
author: "Alice Risely"
date: "16/09/2022"
output:
  html_document:
    toc: yes
    toc_depth: 3
    number_sections: true
    toc_float:
      collapsed: false
    theme: journal
    df_print: paged
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE, message = FALSE, warning = FALSE, fig.height = 5, fig.width =10, time_it = TRUE, error = TRUE)
```

```{r echo = F}
knitr::knit_hooks$set(time_it = local({
  now <- NULL
  function(before, options) {
    if (before) {
      # record the current time before each chunk
      now <<- Sys.time()
    } else {
      # calculate the time difference after a chunk
      res <- difftime(Sys.time(), now, units = "mins")
      # return a character string to show the time
     paste("This chunk took", round(res, 2), "minutes")
    }
  }
}))

```

## Load packages

```{r chunk2, load packages}

library(phyloseq)
library(ggord)
library(ggplot2)
library(viridis)
library(mgcViz)
library(dichromat)
library(ggpattern)
library(expss)
library(dplyr)
library(tidyverse)
library(GGally)
library(vegan)
library(ggpubr)
library(pheatmap)
library(lubridate)
library(ggsci)
library(RColorBrewer)
library(microbiome)
library(vegan)
library(lme4)
library(performance)
library(ggrepel)
library(zoo)

library(glmmTMB)
library(mgcv)
library(gratia)
library(speedyseq)
library(ggeffects)
library(MuMIn)
library(effects)
library(broom)
library(cowplot)
library(patchwork)

library("coxme")
library(survival)

library(SpiecEasi)
library(NetCoMi)

library(gllvm)
library(data.table)
library(datawizard)
library(ggcorrplot)
library(ggthemes)

memory.limit(1000000)

```

# Data Import

```{r}

meerkat_final<-readRDS("C:\\Users\\risel\\Dropbox\\Sommer postdoc\\Meerkat project\\PROJECTS\\MeerkatSpikeInData\\Analysis updated\\6. TB AND MB POPULATION LEVEL\\Published data and code\\V2\\meerkat_MB_GCB.RDS")

daily_rainfall_merged <- read.csv("C:/Users/risel/Dropbox/Sommer postdoc/Meerkat project/PROJECTS/MeerkatSpikeInData/Analysis updated/6. TB AND MB POPULATION LEVEL/Published data and code/V2/rainfall_data.csv", row.names=1)

daily_temp_merged <- read.csv("C:/Users/risel/Dropbox/Sommer postdoc/Meerkat project/PROJECTS/MeerkatSpikeInData/Analysis updated/6. TB AND MB POPULATION LEVEL/Published data and code/V2/maxtemp_data.csv", row.names=1)


TB_stage_df<- read.csv("C:/Users/risel/Dropbox/Sommer postdoc/Meerkat project/PROJECTS/MeerkatSpikeInData/Analysis updated/6. TB AND MB POPULATION LEVEL/Published data and code/V2/TB_data.csv", row.names=1)

daily_rainfall_merged$Date<-as.Date(daily_rainfall_merged$Date, "%Y-%m-%d")
daily_temp_merged$Date<-as.Date(daily_temp_merged$Date, "%Y-%m-%d")
TB_stage_df$Date<-as.Date(TB_stage_df$Date, "%Y-%m-%d")

```

# Count normalisation
* Scale counts by spike in

```{r}

asv_table<-data.frame(meerkat_final@otu_table@.Data)
asv_table[1:5,1:5]
names(asv_table)<-sample_data(meerkat_final)$feature.id
meerkat_normalized<-sweep(asv_table, MARGIN=2, sample_data(meerkat_final)$scaling_factor, `*`)
meerkat_normalized<-round(meerkat_normalized, digits = 0)
meerkat_normalized<-otu_table(meerkat_normalized, taxa_are_rows = TRUE)
meerkat_scaled<-meerkat_final
otu_table(meerkat_scaled)<-meerkat_normalized


### add total abundance of reads to dataframe
sample_data(meerkat_scaled)$TotalAbundance<-sample_sums(meerkat_scaled)

#Filter samples

#meerkat_filtered<-subset_samples(meerkat_scaled, Seq_depth_nospikein >3000)
meerkat_filtered<-subset_samples(meerkat_scaled, sample_sums(meerkat_scaled)>1000)


```

# Fig. 1: Population and sample trends

```{r chunk7,  fig.width = 15, fig.height = 7}

### left panel ###

p1<-ggplot(subset(daily_rainfall_merged, Date > "1996-01-01"), aes(x = Date, y = RollMean12Month_NOAA))+
  geom_point(size =0.1, aes(col = Col))+
  #geom_path(size =0.1)+
  geom_hline(yintercept = mean(daily_rainfall_merged$RollMean12Month_NOAA, na.rm = T), linetype = "longdash")+
 # ggtitle("VZ/KRC data")+
  theme_classic(base_size = 12)+
  ylab("Rainfall (mm)")+
  theme(legend.position = "none")+
  xlim(c(as.Date("1997-01-01"), as.Date("2020-01-01")))+
  theme(panel.grid.major.x = element_line(color = "grey",size = 0.5, linetype = 2))+
 theme(axis.title.x = element_blank())+
  scale_color_manual(values = c("dodgerblue3", "red"))+
  ggtitle("b)")+
  theme(plot.title = element_text( face = "bold", size = 13))



p2<-ggplot(subset(daily_temp_merged, Date > "1996-01-01"), aes(x = Date, y = RollMean12Month_curated))+
  geom_point(size =0.1, aes(col = Col))+
#  geom_path(size =0.1)+
  geom_hline(yintercept = mean(daily_temp_merged$RollMean12Month_curated, na.rm = T), linetype = "longdash")+
 # ggtitle("VZ/KRC data")+
  ylab("Max. temp. (°C)")+
  theme_classic(base_size = 12)+
  theme(legend.position = "none")+
  xlim(c(as.Date("1997-01-01"), as.Date("2020-01-01")))+
  theme(panel.grid.major.x = element_line(color = "grey",size = 0.5, linetype = 2))+
  theme(axis.title.x = element_blank())+
  scale_color_manual(values = c( "red", "dodgerblue3"))+
  ggtitle("a)")+
  theme(plot.title = element_text( face = "bold", size = 13))

p3<-ggplot(TB_stage_df, aes(x = Date, y = Proportion, col = TB_status))+
  geom_point(size = 0.1)+
  geom_path(size = 1)+
  scale_color_manual(values = c("forestgreen", "skyblue","brown2"))+
  theme_classic(base_size = 12)+
  ylab("Proportion population")+
 xlim(c(as.Date("1997-01-01"), as.Date("2020-01-01")))+
  theme(panel.grid.major.x = element_line(color = "grey",size = 0.5, linetype = 2))+
  theme(legend.position = c(0.17,0.5), 
        legend.key.size = unit(0.05, "cm"),
        legend.key.width = unit(0.1, "cm"),
        legend.title = element_text(size = 10) )+
  labs(col = "TB status")+
  ggtitle("c)")+
  theme(plot.title = element_text( face = "bold", size = 13))



Fig1_left<-p2/p1/p3

## right panel ###
## right panel ###


meerkat_class<-tax_glom(meerkat_filtered, taxrank = "Class")

sample_data(meerkat_class)<-sample_data(meerkat_class)[,c("feature.id")]
meerkat_class<-microbiome::transform(meerkat_class, "compositional") 
class_per_sample<-psmelt(meerkat_class)

top_class<-c("Bacteroidia","Fusobacteriia" ,  "Gammaproteobacteria", "Clostridia",  
"Coriobacteriia","Actinobacteria",    "Bacilli")


class_per_sample$Class_plot<-as.character(ifelse(class_per_sample$Class %in% top_class, class_per_sample$Class, "Other"))

class_per_sample$Class_plot<-factor(class_per_sample$Class_plot, levels = c("Bacteroidia","Fusobacteriia" ,  "Other","Gammaproteobacteria", "Clostridia",  
"Coriobacteriia","Actinobacteria",    "Bacilli"))


metadata<-data.frame(sample_data(meerkat_filtered))

metadata<-metadata[,c("feature.id", "SampleDate", "Year", "Storage", "HoursAfterSunrise", "TotalAbundance", "TB_stage", "Season", "Mean_maxtemp_3month",  "Mean_maxtemp_12month","Mean_rainfall_12month_NOAA", "Mean_rainfall_3month_NOAA", "Condition_weight", "GroupAtSampling", "AgeAtSampling")]

class_per_sample<- merge(class_per_sample, metadata, by = "feature.id")

## microbiome plot ######
## microbiome plot ######


pal<-brewer.pal(10,"Paired")
pal<-c(pal, "lightgrey")
pal<-pal[c(2,1,11,4,3,6,9,10)]
pal<-rev(pal)


plot_per_sample<-ggplot(class_per_sample, aes(y = Abundance, x = reorder(Sample, SampleDate), fill = Class_plot))+
  geom_col(position = "fill", stat="identity",  col = NA, width = 2)+ 
  scale_fill_manual(values = rev(pal))+
  ylab("Relative abundance")+
  theme(axis.text.x = element_blank(), axis.ticks = element_blank(), axis.title.x = element_blank())+
   scale_y_continuous(expand = c(0,0)) +
  labs(fill = "Bacterial class")+
   theme(plot.margin=unit(c(0.2,0.2,0.2,1),"cm"))+
  theme(legend.justification = c("left", "center"))



#### temp max ###
#### temp max ###

pal_dichromat<-dichromat::colorschemes$BluetoOrange.12

pal_dichromat[12]<-"black"
pal_dichromat[11]<-"red"


plot_temp1<-ggplot(class_per_sample, aes(y = 1, x = reorder(Sample, SampleDate), fill = Mean_maxtemp_12month))+geom_tile()+
  theme_void()+
 scale_fill_gradientn(colours = pal_dichromat)+
  theme(legend.position="right", legend.direction="vertical", 
         legend.justification = c("left", "top"),
        legend.key.height = unit(0.15, 'cm'),
        legend.title=element_blank(), legend.text=element_text(size=6)) +
  theme(axis.title.y = element_text(angle = 0))+
  ylab(" Max \ntemp")
 
### rainfall ######
### rainfall ######


plot_rainfall2<-ggplot(class_per_sample, aes(y = 1, x = reorder(Sample, SampleDate), fill = Mean_rainfall_12month_NOAA))+geom_tile()+
  theme_void()+
 scale_fill_gradientn(colours = rev(pal_dichromat[1:12]))+
   theme(legend.position="right", legend.direction="vertical", 
        legend.justification = c("left", "top"),
        legend.key.height = unit(0.15, 'cm'),
        legend.title=element_blank(), legend.text=element_text(size=6))+
   theme(axis.title.y = element_text(angle = 0))+
  ylab("Rainfall")


## TB stage ###
## TB stage ###


levels(class_per_sample$TB_stage)
levels(class_per_sample$TB_stage)<-c("TB unexposed", "TB exposed", "TB diseased")
#levels(class_per_sample$TB_stage_weight)<-c("TB unexposed", "TB exposed", "TB diseased")

table(class_per_sample$TB_stage)

plot_TB_sample<-ggplot(class_per_sample, aes(y = 1, x = reorder(Sample, SampleDate), fill = TB_stage))+
  geom_tile()+
  theme_void()+
scale_fill_manual(values = c( "forestgreen", "skyblue","brown2"))+ 
  theme(axis.title.y = element_text(angle = 0))+
  ylab("TB")+
  theme(legend.position="right", legend.direction="vertical", 
       legend.justification = c("left", "top"),
        legend.key.height = unit(0.3, 'cm'),
        legend.title=element_blank())



## season ###
## season ###

plot_season<-ggplot(class_per_sample, aes(y = 1, x = reorder(Sample, SampleDate), fill = Season))+
  geom_tile()+
  theme_void()+
  scale_fill_manual(values = c("gold", "forestgreen"))+
  theme(legend.position="right", legend.direction="vertical", 
        legend.justification = c("left", "top"),
        legend.key.height = unit(0.1, 'cm'),
        legend.title=element_blank())+
   theme(axis.title.y = element_text(angle = 0))+
  ylab("Season")


# year labels ###
# year labels ###

year_names <- c(
                    `1997` = "",
                    `1998` = "",
                    `1999` = "1999",
                    `2000` = "",
                    `2001` = "2001",
                    `2002` = "2002",
                    `2003` = "2003",
                    `2004` = "2004",
                    `2005` = "2005",
                    `2006` = "2006",
                    `2007` = "2007",
                    `2008` = "2008",
                    `2009` = "2009",
                    `2010` = "2010",
                    `2011` = "2011",
                    `2012` = "2012",
                    `2013` = "2013",
                    `2014` = "2014",
                    `2015` = "2015",
                    `2016` = "2016",
                    `2017` = "2017",
                    `2018` = "2018",
                    `2019` = "")

##### year ######
##### year ######
##### year ######

plot_year<-ggplot(class_per_sample, aes(y = 1, x = reorder(Sample, SampleDate)))+
  geom_tile(fill = "skyblue")+
  theme_void()+
 # scale_fill_viridis(discrete = F)+ 
  #scale_fill_manual(values = c("red", "skyblue"))+
  facet_grid(~Year, scales = "free", space = "free_x",switch = "x", labeller = as_labeller(year_names))+

  theme(panel.spacing = unit(0.07, "lines"), 
         strip.background = element_blank(),
         strip.placement = "outside",
         strip.text.x = element_text(angle = 90, hjust = 0.5)) +
  
   xlab("")+
    theme(axis.title.y = element_text(angle = 0, hjust = ))+
  ylab("Year          ")+
  theme(axis.title.x = element_text(colour = "black"))+
  theme(plot.margin=unit(c(0,4.95,0.2,1.25),"cm")) +
    theme(legend.position="right", legend.direction="vertical", 
        legend.justification = c("left", "top"),
        legend.key.height = unit(0.2, 'cm'),
        legend.title=element_blank())

## put it all together


sup_figabc<-ggarrange(plot_per_sample, plot_temp1,plot_rainfall2, plot_TB_sample,  plot_season, 
                      ncol = 1, align = "v", heights = c(10,1,1, 1,  1),
                      labels= c("d)", "e)", "f)", "g)", "h)"),  
                      font.label = list(size = 13))

Fig1_right<-ggarrange(sup_figabc, plot_year, ncol = 1, heights = c(7,1.1), 
                      labels = c(NA, "i)"), font.label = list(size = 13))



# fig 1
ggarrange(Fig1_left, Fig1_right, ncol = 2, widths = c(1,1.8))



```

# Fig 2: Constrained ordination

-   Constrained ordination using CAP
-   Also test all supertypes

```{r chunk16}


scaled_core<-microbiome::core(meerkat_filtered, detection = 100, prevalence = 0.01)
#scaled_core <-microbiome::transform(scaled_core, transform = "clr")


```

-   Generate distance matrices

```{r chunk18}
###### genus level 

wunifrac_dist<-distance(scaled_core, method = "wunifrac")


```


-   organise variables

```{r chunk17}


otutable<-data.frame(t(scaled_core@otu_table@.Data))
dim(otutable)
otutable[1:5,1:5]
colnames(otutable)<-taxa_names(scaled_core)

metadata<-data.frame(sample_data(scaled_core))

time<-metadata$TimeCat
year<-metadata$Year

# methods
storage<-metadata$Storage
seq_depth<-as.numeric(scale(metadata$Seq_depth_nospikein))
seq_run<-metadata$Seq_run


```

* Apply constrained ordination models (dbRDA)

```{r chunk19}


# weighted unifrac

wu_model<-capscale(wunifrac_dist ~ time+
                             year + storage, 
                env = metadata, comm = otutable) 


final_model_anova<-anova.cca(wu_model, by="terms")
final_model_anova



```


* Oridination figure

```{r chunk21}

######################################

## extract data from model

wu_model_df<-scores(wu_model)
str(wu_model_df)


# extract CAP scores

vectors_df<-data.frame(wu_model_df$sites)

vectors_df$feature.id<-row.names(vectors_df)

cap_values_df<-vectors_df

# merge with info on dominant family

sample_metadata<-data.frame(sample_data(scaled_core))
vectors_df<-merge(vectors_df, sample_metadata, by = "feature.id")

sample_data(meerkat_filtered)[1:5,1]
sample_data(meerkat_final)[1:5,1]
cap_values_df[1:5,]
sample_data(meerkat_filtered)[20:25,1]
sample_data(meerkat_final)[20:25,1]
cap_values_df[20:25,]


sample_data(meerkat_filtered)$CAP1<-cap_values_df$CAP1
sample_data(meerkat_filtered)$CAP2<-cap_values_df$CAP2

#########################

p_beta1<-ggplot(vectors_df, aes(x = CAP1, y = CAP2))+
  stat_ellipse(geom = "polygon", aes(fill = YearCat, col = YearCat), level = 0.9, alpha = 0.2, size = 0.5)+
  geom_point(aes(fill =YearCat), pch = 21, size = 1.5, col = "black")+
  theme_classic(base_size = 12)+
   scale_fill_viridis(discrete = TRUE, direction = -1)+scale_color_viridis(discrete = TRUE, direction = -1)+
  labs(fill = "Period", col = "Period")+
 geom_point(data = subset(vectors_df, YearCat == "Pre-2005"), pch = 21, size = 1.5, col = "black", fill = "yellow")+
  theme(legend.position = c(0.85,0.9))+
  theme(legend.key.size = unit(0.2, "cm"),
        legend.title = element_blank())


vectors_df$TimeCat<-ifelse(vectors_df$TimeCat == "MORNING", "Morning", "Afternoon")
vectors_df$TimeCat<-factor(vectors_df$TimeCat, levels = c("Morning", "Afternoon"))

p_beta3<-ggplot(vectors_df, aes(x = CAP1, y = CAP2))+
  stat_ellipse(geom = "polygon", aes(fill = TimeCat, col = TimeCat), level = 0.9, alpha = 0.2, size = 0.5)+
  geom_point(aes(fill =TimeCat), pch = 21, size = 1.5,  col = "black")+
  theme_classic(base_size = 12)+
   scale_fill_manual(values = c("skyblue", "red"))+
   scale_color_manual(values = c( "skyblue", "red"))+
  labs(fill = "Time", col = "Time")+
  theme(legend.position = c(0.9,0.9))+
  theme(legend.key.size = unit(0.2, "cm"),
        legend.title = element_blank())



p_beta2<-ggplot(vectors_df, aes(x = YearCat, y = CAP2, fill = YearCat, col = YearCat))+
  geom_jitter(alpha =0.4, width = 0.2,  pch = 21, size = 0.5)+
  geom_boxplot( pch = 21, size = 0.5, col = "black")+
   scale_fill_viridis(discrete = TRUE, direction = -1)+
 scale_color_viridis(discrete = TRUE, direction = -1)+
  theme_classic(base_size = 12)+
  labs(fill = "Period", col = "Period")+
    theme(axis.text.x = element_text(angle = 45, vjust = 1, hjust=1))+
  theme(axis.title.x = element_blank())+
theme(legend.position = "none")


# model CAP2 scores 

beta_model<-lmer(CAP2~YearCat +(1|IndividID), data = vectors_df) #gaussian

summary(beta_model)
sjPlot::tab_model(beta_model)

# ICC value 
icc(beta_model)


```


* Microbiome composition by year barplot

```{r chunk15}


meerkat_merged_year<-merge_samples(meerkat_filtered,"Year", fun=mean)

sample_data(meerkat_merged_year)$Year<-sample_names(meerkat_merged_year)

meerkat_merged_year_class<-tax_glom(meerkat_merged_year, taxrank = "Class")

sample_data(meerkat_merged_year_class)<-sample_data(meerkat_merged_year_class)[,1]

class_per_year<-psmelt(meerkat_merged_year_class)


class_per_year$Class_plot<-as.character(ifelse(class_per_year$Class %in% top_class, class_per_year$Class, "Other"))

class_per_year$Class_plot<-factor(class_per_year$Class_plot, levels = c("Bacteroidia","Fusobacteriia" ,  "Other","Gammaproteobacteria", "Clostridia",  
"Coriobacteriia","Actinobacteria",    "Bacilli"))

class_per_year$Sample<- as.integer(class_per_year$Sample)

class_per_year<-class_per_year%>% group_by(Sample, Class_plot)%>%summarize(Abundance = sum(Abundance))
class_per_year<-data.frame(class_per_year)
class_per_year$Sample<-factor(class_per_year$Sample)

plot_mb_year<-ggplot(class_per_year, aes(x =Sample, y = Abundance, fill = Class_plot))+
  geom_bar( stat="identity",  size = 0, position = "fill")+ 
   theme_classic(base_size = 12)+
  scale_fill_manual(values = rev(pal))+
  theme(legend.position = "right")+
  xlab("")+ 
  theme(axis.text.x = element_text(angle = 90, vjust = 0.9, hjust=1))+
     theme(axis.text.y = element_text(size=8))+
  labs(fill = "Bacterial class")+
   scale_y_continuous(expand = c(0,0))+
  xlab("Year")+
  theme(legend.text = element_text(size=8),
  legend.title = element_text(size=10))+
  theme(legend.key.height = unit(0.3,'cm'))+
  theme(legend.key.width = unit(0.3,'cm'))+
  theme(axis.text.x = element_text(size = 8))+
  theme(plot.margin=unit(c(0,0.2,0.2,0.2),"cm"))

##############
##############
##############

year_freq<-data.frame(table(sample_data(meerkat_filtered)$Year))
names(year_freq)[1]<-"Year"

year_freq$Year<-as.numeric(as.character(year_freq$Year))

year_freq<-year_freq %>% mutate(YearCat = case_when((Year <2005 ~ "Pre-2005"),
                                                  (Year >=2005 & Year < 2010) ~ "2005-2010",
                                                  (Year >=2010 & Year < 2015) ~ "2010-2015",
                                                  (Year >=2015) ~ "2015-2019"))

year_freq$YearCat<-factor(year_freq$YearCat, levels = c("Pre-2005", "2005-2010", "2010-2015", "2015-2019"))


year_freq$Year<-factor(year_freq$Year)

#######
#######
#######


year_hist<-ggplot(year_freq, aes(x = Year, y = Freq, fill = YearCat)) + 
   geom_bar( stat = "identity", size = 0)+ 
# theme(plot.margin=unit(c(0.2,5,0.2,0.2),"cm"))+
  theme(axis.text.x = element_blank(), axis.title.x = element_blank())+
  theme(legend.position = "right")+
  scale_y_sqrt(breaks = c(0,20,50,100,200))+
 # geom_hline(yintercept = 20, col = "black", linetype = "longdash")+
     scale_fill_viridis(discrete = TRUE, direction = -1)+
 scale_color_viridis(discrete = TRUE, direction = -1)+
    theme(legend.key.height = unit(0.1, 'cm'),
          legend.title = element_blank())+
   theme(legend.text = element_text(size=8))+
   theme(axis.text.y = element_text(size=6))+
  theme(plot.margin=unit(c(0.2,0.2,0,0.2),"cm"))

year_plot<-ggarrange(year_hist, plot_mb_year, ncol = 1, align = "v", heights = c(1,4))+theme(plot.margin=unit(c(0.2,0.2,0.2,0.4),"cm"))


```

* Ordination by individual ID

```{r, fig.width = 10, fig.height = 6, }

meerkat_merged_ID<-merge_samples(meerkat_filtered,"IndividID", fun=mean)
sample_data(meerkat_merged_ID)$IndividID<-sample_names(meerkat_merged_ID)
sample_data(meerkat_merged_ID)<-sample_data(meerkat_merged_ID)[,2]

scaled_core_id<-microbiome::core(meerkat_merged_ID, detection = 100, prevalence = 0.01)

wunifrac_dist_id<-distance(scaled_core_id, method = "wunifrac")

############

otutable<-data.frame(scaled_core_id@otu_table@.Data)
dim(otutable)
otutable[1:5,1:5]
colnames(otutable)<-taxa_names(scaled_core_id)

metadata<-data.frame(sample_data(scaled_core_id))
metadata_original<- data.frame(sample_data(meerkat_filtered))
metadata$BirthDate<-vlookup(metadata$IndividID, metadata_original, lookup_column = "IndividID", result_column = "BirthDate")
metadata$BirthYear<-lubridate::year(metadata$BirthDate)

metadata<-metadata%>% mutate(YearCat = case_when((BirthYear <2005 ~ "Pre-2005"),
                                                  (BirthYear >=2005 & BirthYear < 2010) ~ "2005-2010",
                                                  (BirthYear >=2010 & BirthYear < 2015) ~ "2010-2015",
                                                  (BirthYear >=2015) ~ "2015-2019"))

metadata$YearCat<-factor(metadata$YearCat, levels = c("Pre-2005", "2005-2010", "2010-2015", "2015-2019"))


# year


YearCat<-as.character(metadata$YearCat)
year<-metadata$BirthYear
table(metadata$YearCat)

id_model<-capscale(wunifrac_dist_id ~ 
                             year, 
                env = metadata, comm = otutable) 

## extract data from model

final_model_df<-scores(id_model)

# extract CAP scores

vectors_df_id<-data.frame(final_model_df$sites)

vectors_df_id$feature.id<-row.names(vectors_df_id)

# merge with info on dominant family

vectors_df_id<-merge(vectors_df_id, metadata, by.x = "feature.id", by.y = "IndividID")

############
############
############


#########################



beta_id<-ggplot(vectors_df_id, aes(x = YearCat, y = CAP1, fill = YearCat, col = YearCat))+
  geom_jitter(alpha =0.4, width = 0.2,  pch = 21, size = 0.5)+
  geom_boxplot( pch = 21, size = 0.5, col = "black")+
   scale_fill_viridis(discrete = TRUE, direction = -1)+
 scale_color_viridis(discrete = TRUE, direction = -1)+
  theme_classic(base_size = 12)+
  labs(fill = "Period", col = "Period")+
    theme(axis.text.x = element_text(angle = 45, vjust = 1, hjust=1))+
  theme(axis.title.x = element_blank())+
theme(legend.position = "none")



p1<-ggarrange(year_plot, p_beta1, ncol = 2,  labels = c("a)", "b)"))
p2<-ggarrange(p_beta3, p_beta2, beta_id, ncol = 3,  widths = c(2,1,1), labels = c("c)", "d)", "e)"))

ggarrange(p1, p2, nrow = 2)


```


# Fig. 3: GAMs and non-linear trends

```{r chunk22}


meerkat_genus<-microbiome::aggregate_top_taxa(meerkat_filtered, "Genus", top = 50)
clr_scaled <-microbiome::transform(meerkat_genus, transform = "clr")


taxa_to_keep <- c("[Ruminococcus] torques group" , "Arthrobacter", "Bacillus","Bacteroides",
 "Blautia", "Catenisphaera" ,"Christensenellaceae R-7 group", "Clostridium sensu stricto 1",
"Collinsella" ,"Enterococcus" , "Escherichia-Shigella","Fusobacterium",
"Geodermatophilus","Lachnoclostridium", "Lactococcus" ,"Marvinbryantia" ,
"Paeniclostridium","Peptoclostridium" , "Peptococcus" ,"Phascolarctobacterium",
"Romboutsia", "Slackia" ,"Turicibacter" , "Weissella" , 
"Rikenellaceae RC9 gut group", "Helicobacter" , "Parabacteroides", "Lactobacillus",
 "Pediococcus", "Streptococcus" )

#meerkat_genus<-microbiome::aggregate_top_taxa(meerkat_filtered, "Genus", top = 50)

#meerkat_genus_scaled<-prune_taxa(taxa_to_keep, meerkat_genus)
meerkat_genus_scaled<-prune_taxa(taxa_to_keep,clr_scaled) # new

genus_ps<-meerkat_genus_scaled
sample_data(genus_ps)<-sample_data(genus_ps)[,c( "feature.id")]

top_genera_melt<-psmelt(genus_ps)
head(top_genera_melt)

metadata<-data.frame(sample_data(meerkat_genus_scaled))

top_genera_melt3<- merge(top_genera_melt, metadata, by.x = "Sample", by.y = "feature.id")
names(top_genera_melt3)

head(top_genera_melt3)


```


```{r chunk23}

# loop to fit GAMMs to each genus

year_estimates<-list()

taxanames<-unique(top_genera_melt3$Genus)

for (i in 1:length(taxanames)){
  tryCatch({ #catch errors
    print(i)
    print(taxanames[i])
    
    taxa1<-subset(top_genera_melt3, Genus == taxanames[i])
    # taxa1<-subset(top_genera_melt3, Genus == "Bacteroides")
    
    metadata<-taxa1

  # scale variables
    
    metadata$Time_scaled<-as.numeric(rescale(metadata$HoursAfterSunrise, to = c(-1,1)))
    

### seq depth 

metadata$Seq_depth_scaled <- as.numeric(rescale(metadata$Seq_depth_nospikein, to = c(-1,1)))

metadata$Date<- as.numeric(metadata$SampleDate - min(metadata$SampleDate))
metadata$Date_scaled<-as.numeric(rescale(metadata$Date, to = c(-1,1)))

metadata$Maxtemp_12month_scaled <-as.numeric(rescale(metadata$Mean_maxtemp_12month, to = c(-1,1)))


metadata$Age_scaled<-as.numeric(rescale(metadata$AgeAtSampling, to = c(-1,1)))


    
    # fit gamm    
    m_taxa <- mgcv::bam(Abundance~
                          s(Date_scaled, bs = "cr", k = 6)+ # specifically interested in this variable
                          s(Time_scaled, bs = "cr")+
                          s(month, bs = "cc")+
                          s(Age_scaled, bs = "cr")+
                          s(Seq_depth_scaled, bs = "cr")+
                          Storage+
                          Season+
                          s(IndividID, bs = "re")+
                          s(GroupAtSampling, bs = "re"),
                        #correlation = corARMA(form = ~ 1|Year, p = 3),
                        data=metadata,
                        family = gaussian)
    
    print(summary(m_taxa))
    #gam.check(m_taxa)
    #plot(m_taxa)
    
    
    
    confint_df_year<-stats::confint(m_taxa, parm = "s(Date_scaled)",  type = "confidence")
    confint_df_year$Taxa<-taxanames[i]
    confint_df<-confint_df_year
    
    # convert into real estimate by adding the intercapt (= mean abundance)
    
    confint_df$est_real<-confint_df$est+summary(m_taxa)$p.coeff[1] 
    confint_df$lower_real<- confint_df$lower+summary(m_taxa)$p.coeff[1] 
    confint_df$upper_real<- confint_df$upper+summary(m_taxa)$p.coeff[1] 
    
    ## add effect size and p value
    
    summary<-summary(m_taxa)
    confint_df$effect_size<-  summary$s.table[1,3]
    confint_df$P_val<-  summary$s.table[1,4]
    
    year_estimates[[i]]<-confint_df
    
    
    
  }, error=function(e){})
}



names(year_estimates)<-taxanames

year_estimates_df<-data.frame(do.call(rbind, year_estimates))



```

```{r}
year_estimates_df$Significant <-year_estimates_df$P_val<0.05

year_estimates_df$Year<-as.numeric(rescale(year_estimates_df$Date_scaled, to = c(1997,2019)))


ggplot(year_estimates_df, aes(x = Year, y = est))+geom_line()+geom_ribbon(aes(ymin = lower, ymax = upper, fill = Significant), alpha = 0.5)+
  facet_wrap(~Taxa)+ theme_bw()+
  scale_fill_manual(values = c("grey", "navyblue"))+
  theme(axis.text.x = element_text(angle = 45, vjust=0.7))+
  geom_hline(yintercept = 0, linetype = "dashed")+
  coord_cartesian(ylim = c(-3,3))




year_estimates_red<- subset(year_estimates_df, Taxa == "Lactococcus"| Taxa == "Pediococcus"| Taxa == "Weissella"|Taxa == "Bacteroides" | Taxa == "Fusobacterium" | Taxa == "Rikenellaceae RC9 gut group" )



year_estimates_red$Taxa<-factor(year_estimates_red$Taxa, levels = c("Bacteroides",  "Rikenellaceae RC9 gut group","Fusobacterium", "Lactococcus", "Pediococcus", "Weissella"))
unique(year_estimates_red$Taxa)

year_estimates_red$Class <- ifelse(year_estimates_red$Taxa == "Bacteroides" | year_estimates_red$Taxa == "Rikenellaceae RC9 gut group", "Bacteroidia", NA)
year_estimates_red$Class <- ifelse(year_estimates_red$Taxa == "Fusobacterium" , "Fusobacteria", year_estimates_red$Class)
year_estimates_red$Class <- ifelse(year_estimates_red$Taxa == "Lactococcus" | year_estimates_red$Taxa == "Pediococcus"| year_estimates_red$Taxa == "Weissella", "Bacilli",  year_estimates_red$Class)

levels(year_estimates_red$Taxa)<-c("a) Bacteroides", "b) Rikenellaceae RC9","c) Fusobacterium", "d) Lactococcus (LAB)", "e) Pediococcus (LAB)", "f) Weissella (LAB)")



ggplot(year_estimates_red,aes(x = Year, y = est))+
  geom_line(size = 0.8, aes(col = Class))+
  geom_line(size = 0.8, alpha = 0.4, col = "black")+
  facet_wrap(~Taxa, ncol = 3)+
  geom_ribbon(aes(ymin = lower, ymax = upper, fill = Class ), alpha = 0.7)+
  coord_cartesian(ylim = c(-3.5,3.5))+
  geom_hline(yintercept = 0, linetype = "dashed")+
  scale_fill_manual(values = c(pal[1], pal[8], pal[7]))+
  scale_color_manual(values = c(pal[1], pal[8], pal[7]))+
   theme_fira() +
 theme(plot.margin=unit(c(0.2,0.5,0,0.2),"cm"))+
  #labs(fill = "Class", col = "Genus")+
  ylab("Estimate")+
  xlab("")+
    theme(axis.text.x = element_text(angle = 45, vjust=0.7))+
 annotate("segment", x=-Inf, xend=Inf, y=-Inf, yend=-Inf, col = "black")+
  annotate("segment", x=-Inf, xend=-Inf, y=-Inf, yend=Inf, col = "darkgrey")+
  theme(panel.grid.major = element_blank())


# firatheme::firaSave("C:/Users/risel/Dropbox/Sommer postdoc/Meerkat project/PROJECTS/MeerkatSpikeInData/Manuscripts/TB population level paper/GCB/Resubmission/Figures/fig3.pdf", device = "pdf")


```



# Fig. 4: JDSMs

```{r}

# Using scaled counts


meerkat_genus<-microbiome::aggregate_top_taxa(meerkat_filtered, "Genus", top = 50)
clr_scaled <-microbiome::transform(meerkat_genus, transform = "clr")


taxa_to_keep <- c("[Ruminococcus] torques group" , "Arthrobacter", "Bacillus","Bacteroides",
 "Blautia", "Catenisphaera" ,"Christensenellaceae R-7 group", "Clostridium sensu stricto 1",
"Collinsella" ,"Enterococcus" , "Escherichia-Shigella","Fusobacterium",
"Geodermatophilus","Lachnoclostridium", "Lactococcus" ,"Marvinbryantia" ,
"Paeniclostridium","Peptoclostridium" , "Peptococcus" ,"Phascolarctobacterium",
"Romboutsia", "Slackia" ,"Turicibacter" , "Weissella" , 
"Rikenellaceae RC9 gut group", "Helicobacter" , "Parabacteroides", "Lactobacillus",
 "Pediococcus", "Streptococcus" )

#meerkat_genus<-microbiome::aggregate_top_taxa(meerkat_filtered, "Genus", top = 50)

#meerkat_genus_scaled<-prune_taxa(taxa_to_keep, meerkat_genus)
meerkat_genus_scaled<-prune_taxa(taxa_to_keep,clr_scaled) # new
meerkat_genus_scaled<-subset_samples(meerkat_genus_scaled, !is.na(Mean_rainfall_12month_NOAA))


#############

taxa_names(meerkat_genus_scaled)
taxa_names(meerkat_genus_scaled)[10]<-paste(taxa_names(meerkat_genus_scaled)[10], "(LAB)")
taxa_names(meerkat_genus_scaled)[16]<-paste(taxa_names(meerkat_genus_scaled)[16], "(LAB)")
taxa_names(meerkat_genus_scaled)[17]<-paste(taxa_names(meerkat_genus_scaled)[17], "(LAB)")
taxa_names(meerkat_genus_scaled)[21]<-paste(taxa_names(meerkat_genus_scaled)[21], "(LAB)")
taxa_names(meerkat_genus_scaled)[28]<-paste(taxa_names(meerkat_genus_scaled)[28], "(LAB)")
taxa_names(meerkat_genus_scaled)[30]<-paste(taxa_names(meerkat_genus_scaled)[30], "(LAB)")

```

* Data preparation

```{r}

ys <- data.frame(t(otu_table(meerkat_genus_scaled)))
  
names(ys) <-taxa_names(meerkat_genus_scaled)


#y<-data.frame(scale(y))
Xs<-data.frame(sample_data(meerkat_genus_scaled))



Xs$Rainfall_1month_scaled <-as.numeric(rescale(Xs$Mean_rainfall_1month_NOAA, to = c(-1,1)))
hist(Xs$Rainfall_1month_scaled)

Xs$Rainfall_3month_scaled <-as.numeric(rescale(Xs$Mean_rainfall_3month_NOAA, to = c(-1,1)))
hist(Xs$Rainfall_3month_scaled)

Xs$Rainfall_6month_scaled <-as.numeric(rescale(Xs$Mean_rainfall_6month_NOAA, to = c(-1,1)))
hist(Xs$Rainfall_6month_scaled)

Xs$Rainfall_12month_scaled <-as.numeric(rescale(Xs$Mean_rainfall_12month_NOAA, to = c(-1,1)))
hist(Xs$Rainfall_12month_scaled)

## KRC/VZ rainfall data 

Xs$Rainfall_3month_scaled_local <-as.numeric(rescale(Xs$Mean_rainfall_3month, to = c(-1,1)))
hist(Xs$Rainfall_3month_scaled_local)

Xs$Rainfall_12month_scaled_local <-as.numeric(rescale(Xs$Mean_rainfall_12month, to = c(-1,1)))
hist(Xs$Rainfall_12month_scaled_local)


## temperature


Xs$Maxtemp_1month_scaled <-as.numeric(rescale(Xs$Mean_maxtemp_1month, to = c(-1,1)))
hist(Xs$Maxtemp_1month_scaled)

Xs$Maxtemp_3month_scaled <-as.numeric(rescale(Xs$Mean_maxtemp_3month, to = c(-1,1)))
hist(Xs$Maxtemp_3month_scaled)

Xs$Maxtemp_6month_scaled <-as.numeric(rescale(Xs$Mean_maxtemp_6month, to = c(-1,1)))
hist(Xs$Maxtemp_6month_scaled)

Xs$Maxtemp_12month_scaled <-as.numeric(rescale(Xs$Mean_maxtemp_12month, to = c(-1,1)))
hist(Xs$Maxtemp_12month_scaled)


### seq depth 
### seq depth 

#Xs$Abundance_scaled <- as.numeric(rescale(Xs$TotalAbundance, to = c(-1,1)))
Xs$Seq_depth_scaled <- as.numeric(rescale(Xs$Seq_depth_nospikein, to = c(-1,1)))
hist(Xs$Seq_depth_scaled)


#Xs$TotalAbundance_scaled <- rescale(Xs$TotalAbundance, to = c(-1,1))

Xs$Year<-as.factor(Xs$Year)

min(Xs$SampleDate)
Xs$Date<- as.numeric(Xs$SampleDate - min(Xs$SampleDate))
Xs$Date_scaled<-as.numeric(rescale(Xs$Date, to = c(-1,1)))


Xs$Condition_scaled<-as.numeric(rescale(Xs$Condition_weight, to = c(-1,1)))
Xs$Age_scaled<-as.numeric(rescale(Xs$AgeAtSampling, to = c(-1,1)))

Xs$TB_stage<-factor(Xs$TB_stage, levels = c( "Unexposed","Exposed", "Diseased"))

names(Xs)

Xs<-Xs[,c(1,2, 15, 5, 22, 24, 21, 31,33, 49:60,62:64)]
names(Xs)


mr<-mean(Xs$Rainfall_12month_scaled)
mt<-mean(Xs$Maxtemp_12month_scaled)

Xs$Climate_cat <-ifelse(Xs$Maxtemp_12month_scaled>mt & Xs$Rainfall_12month_scaled < mr, "HighTemp_LowRainfall", NA)
Xs$Climate_cat <-ifelse(Xs$Maxtemp_12month_scaled<mt & Xs$Rainfall_12month_scaled < mr, "LowTemp_LowRainfall", Xs$Climate_cat)
Xs$Climate_cat <-ifelse(Xs$Maxtemp_12month_scaled<mt & Xs$Rainfall_12month_scaled > mr, "LowTemp_HighRainfall", Xs$Climate_cat)
Xs$Climate_cat <-ifelse(Xs$Maxtemp_12month_scaled>mt & Xs$Rainfall_12month_scaled > mr, "HighTemp_HighRainfall", Xs$Climate_cat)
table(Xs$Climate_cat)

Xs$Climate_cat<-factor(Xs$Climate_cat, levels = c("LowTemp_HighRainfall", "HighTemp_LowRainfall", "HighTemp_HighRainfall", "LowTemp_LowRainfall"))

Xs$Season <-factor(Xs$Season, levels = c("WET", "DRY"))



head(ys)[1:5,1:5]

ggpairs(Xs[,c(6,11,17,13,19)])


```


* Model fitting - reduced model with just date and methodological variables (plus time of day - biggest driver of composition)

```{r}

fit_reduced_scaled <- gllvm(ys, Xs, 
             num.lv = 2,
            formula = ~  Date_scaled+Storage+TimeCat+ Seq_depth_scaled, 
             family = "gaussian",
            row.eff = ~(1|IndividID)+ (1|GroupAtSampling))

coefplot(fit_reduced_scaled, cex.ylab = 0.5 ,mar = c(5,12,2,1), which.Xcoef = c(1))


```


* Extract estimates

```{r}

df<-coef(fit_reduced_scaled)
est_df<-data.frame(df$Intercept)
est_df2<-data.frame(df$Xcoef) 
est_df3<-merge(est_df, est_df2, by = 0)

# order genera

row.names(est_df3)<-est_df3$Row.names
est_df3<-est_df3[colnames(ys),]

#put est_df3 into long format

names(est_df3)[1]<- "Genus"
names(est_df3)[2]<- "Intercept"

estimates_df <- gather(est_df3, Treatment, Estimate, names(est_df3)[2]:names(est_df3)[ncol(est_df3)], factor_key=TRUE)

############# extract confindence intervals  ####
############# extract confindence intervals  ####
############# extract confindence intervals  ####
############# extract confindence intervals  ####


confint_df<-data.frame(confint(fit_reduced_scaled))


confint_df<-rbind(confint_df[rownames(confint_df) %like% "Xcoef", ],
confint_df[rownames(confint_df) %like% "Intercept", ])

head(confint_df)

# add a column with correct variable level
variables<- colnames(est_df3)[3:ncol(est_df3)]
variables<-c(variables, "Intercept")
variables1<-rep(variables, nrow(est_df))
variables2<-variables1[order(match(variables1, variables))]

confint_df$Treatment<-variables2


# column with taxa names. Should be automatically in the correct order but double check

confint_df$Genus<-rep(colnames(ys), length(unique(confint_df$Treatment)))


merged<-merge(estimates_df, confint_df, by = c("Treatment", "Genus"))


final_estimates_reduced<- merged
names(final_estimates_reduced)[4]<-"CI_lower"
names(final_estimates_reduced)[5]<-"CI_upper"

#final_estimates<- merged[,c(1,2,3,7,8)]
head(final_estimates_reduced)
unique(final_estimates_reduced$Treatment)

final_estimates_reduced2<-subset(final_estimates_reduced, Treatment == "Date_scaled")

## add significance

final_estimates_reduced2$Sig<- !data.table::between(0, final_estimates_reduced2$CI_lower, final_estimates_reduced2$CI_upper)


final_estimates_reduced2$Genus<-factor(final_estimates_reduced2$Genus)

genus_order<-as.character((final_estimates_reduced2 %>% arrange(Estimate))$Genus)


```

*full model

* see JDSM_model_selection.R for model selection 

```{r}


########## final model

m_final <- gllvm(ys, Xs, 
             num.lv = 2,
             formula = ~ 
             TB_stage + 
             Season+
             Maxtemp_12month_scaled+ 
             Seq_depth_scaled+
             Storage + 
             TimeCat+
             Age_scaled+
             Condition_scaled, 
             family = "gaussian",
             row.eff = ~(1|IndividID)+ (1|GroupAtSampling))


coefplot(m_final, which.Xcoef = c(1:4, 8), cex.ylab = 0.7)

### explanatory power ###


linpred_df<- data.frame(residuals(m_final)$linpred)
head(linpred_df)
names(linpred_df)<-names(ys)


r2_list<-list()
taxanames<-names(linpred_df)

for (i in 1:length(taxanames)){
r <- as.numeric(cor(linpred_df[,i], ys[i]))
r2<-r^2

r2_list[i]<- r2
}

r2_df<-data.frame(do.call(rbind, r2_list))
r2_df$Genus<-taxanames
names(r2_df)[1]<-"R2"
summary(r2_df$R2)
r2_df %>% arrange(R2)

# Min.   1st Qu.  Median    Mean 3rd Qu.    Max. 
#0.07585 0.23246 0.33921 0.38398 0.54571 0.76018 

#####


m_climate <- gllvm(ys, Xs, 
             num.lv = 2,
             formula = ~ 
             TB_stage + 
            Season+
             Climate_cat+
             Seq_depth_scaled+
             Storage + 
             TimeCat+
             Age_scaled+
            Condition_scaled, 
             family = "gaussian",
             row.eff = ~(1|IndividID)+ (1|GroupAtSampling))
coefplot(m_climate, which.Xcoef = c(3:6, 11), cex.ylab = 0.7)


#### null model 

m_null2 <- gllvm(ys, Xs, 
             num.lv = 2,
             formula = ~ 
            # TB_stage + 
            #Season+
             #Rainfall_3month_scaled + 
             #Rainfall_12month_scaled+
            # Maxtemp_3month_scaled+
             #Maxtemp_12month_scaled+ 
             Seq_depth_scaled+
             Storage+
             TimeCat+
             Age_scaled,
            #Condition_scaled, 
             family = "gaussian",
            row.eff = ~(1|IndividID)+ (1|GroupAtSampling))

cov_env<-getResidualCov(m_final, adjust = 0)#
cov_null<-getResidualCov(m_null2, adjust = 0)#

# variance explained (full model compared to null model)
1 - cov_env$trace / cov_null$trace


################
################
################
################

# effect of rainfall

m_rainfall <- gllvm(ys, Xs, 
             num.lv = 2,
             formula = ~ 
             TB_stage + 
              Season+
             Rainfall_12month_scaled+
             Seq_depth_scaled+
             Storage + 
             TimeCat+
             Age_scaled+
            Condition_scaled, 
             family = "gaussian",
             row.eff = ~(1|IndividID)+ (1|GroupAtSampling))

coefplot(m_rainfall, which.Xcoef = c(4), cex.ylab = 0.7, mar = c(5,12,2,1))

AIC(m_rainfall, m_final) # AIC comparison


```

*Extract estimates 

```{r}
genus_order<-as.character((final_estimates_reduced2 %>% arrange(Estimate))$Genus)


df<-coef(m_final)
est_df<-data.frame(df$Intercept)
est_df2<-data.frame(df$Xcoef) # choose columns of interest
est_df3<-merge(est_df, est_df2, by = 0)


# order genera

row.names(est_df3)<-est_df3$Row.names
est_df3<-est_df3[colnames(ys),]

#put est_df3 into long format

names(est_df3)[1]<- "Genus"
names(est_df3)[2]<- "Intercept"

estimates_df <- gather(est_df3, Treatment, Estimate, names(est_df3)[2]:names(est_df3)[ncol(est_df3)], factor_key=TRUE)

############# extract confindence intervals  ####
############# extract confindence intervals  ####
############# extract confindence intervals  ####
############# extract confindence intervals  ####


confint_df<-data.frame(confint(m_final))
confint_df<-rbind(confint_df[rownames(confint_df) %like% "Xcoef", ],
confint_df[rownames(confint_df) %like% "Intercept", ])


# add a column with correct variable level
variables<- colnames(est_df3)[3:ncol(est_df3)]
variables<-c(variables, "Intercept")
variables1<-rep(variables, nrow(est_df))
variables2<-variables1[order(match(variables1, variables))]


confint_df$Treatment<-variables2


# column with taxa names. Should be automatically in the correct order but double check

confint_df$Genus<-rep(colnames(ys), length(unique(confint_df$Treatment)))

# now have estimates, confidence intervals, aas seperate data frames, but they are in different formats. Need to massage them into one dataframe for plotting.

merged<-merge(estimates_df, confint_df, by = c("Treatment", "Genus"))


final_estimates<- merged
names(final_estimates)[4]<-"CI_lower"
names(final_estimates)[5]<-"CI_upper"


final_estimates2<-subset(final_estimates, Treatment == "TB_stageExposed" | Treatment == "TB_stageDiseased"| Treatment ==  "SeasonDRY"| Treatment ==  "Maxtemp_12month_scaled" |Treatment ==  "Condition_scaled")
unique(final_estimates2$Treatment)


## add significance

final_estimates2$Sig<- !data.table::between(0, final_estimates2$CI_lower, final_estimates2$CI_upper)


#########
#########
#########
#########

final_estimates3<-rbind(final_estimates2, final_estimates_reduced2)

levels(final_estimates3$Treatment)
final_estimates3$Treatment<-factor(final_estimates3$Treatment)

final_estimates3$Treatment<-factor(final_estimates3$Treatment, levels = c( "Date_scaled","SeasonDRY", "Maxtemp_12month_scaled", "TB_stageExposed",  "TB_stageDiseased" , "Condition_scaled"))


### tax table ###
### tax table ###
### tax table ###
### tax table ###

meerkat_genus<-tax_glom(meerkat_final, taxrank = "Genus")
meerkat_genus<-core(meerkat_genus, detection = 0, prev = 0.1)

taxonomy<-data.frame(tax_table(meerkat_genus))
tax<-taxonomy[,c("Genus", "Class")]
tax<-distinct(tax, Genus, .keep_all = T)

unique(tax$Genus)%>%sort()
unique(final_estimates3$Genus) %>% sort()

tax$Genus <- ifelse(tax$Genus == "Lactococcus", "Lactococcus (LAB)", tax$Genus)
tax$Genus <- ifelse(tax$Genus == "Lactobacillus", "Lactobacillus (LAB)", tax$Genus)
tax$Genus <- ifelse(tax$Genus == "Enterococcus", "Enterococcus (LAB)", tax$Genus)
tax$Genus <- ifelse(tax$Genus == "Pediococcus", "Pediococcus (LAB)", tax$Genus)
tax$Genus <- ifelse(tax$Genus == "Streptococcus", "Streptococcus (LAB)", tax$Genus)
tax$Genus <- ifelse(tax$Genus == "Weissella", "Weissella (LAB)", tax$Genus)


final_estimates3$Class<-expss::vlookup(final_estimates3$Genus, tax, lookup_column = "Genus", result_column = "Class")


final_estimates3$Genus<-factor(final_estimates3$Genus, levels = genus_order)

# code genus colours

pal<-brewer.pal(10,"Paired")
pal<-c(pal, "lightgrey")
pal<-pal[c(2,1,11,4,3,6,9,10)]
pal<-rev(pal)
scales::show_col(pal)

final_estimates3 <- final_estimates3 %>% 
  mutate(Colour = case_when((Class == "Bacteroidia") ~ pal[8],
                             (Class == "Fusobacteriia") ~ pal[7],
                             (Class == "Clostridia") ~ pal[5],
                             (Class == "Bacilli") ~ pal[1] )) 

final_estimates3$Colour<-ifelse(is.na(final_estimates3$Colour), "black", as.character(final_estimates3$Colour))


final_estimates3<-final_estimates3 %>% arrange( Treatment, Genus)

# variable labels

variable_names <- c(
                    `Date_scaled` = "a) Date",
                    `TB_stageExposed` = "d) TB status \n(Exposed)",
                    `TB_stageDiseased` = "e) TB status \n(Diseased)",
                    `SeasonDRY` = "b) Season \n(Dry)",
                    `Maxtemp_12month_scaled` = "c) Long-term \nmax. temp.",
                    `Condition_scaled` = "f) Body \ncondition"
                    )


final_estimates3<-final_estimates3 %>% arrange( Treatment, Genus)

## dummy data to keep x axis of most variables constant

Treatment <- rep(as.character(unique(final_estimates3$Treatment)),2)
Estimate<- c(2,2,2,2,2,2,-2,-2,-2,-2,-2,-2)
Genus<- c("Bacillus","Bacillus", "Bacillus","Bacillus","Bacillus", "Bacillus","Collinsella","Collinsella", "Collinsella","Collinsella","Collinsella", "Collinsella" )
Sig<-c(TRUE, TRUE, TRUE,TRUE, TRUE, TRUE,TRUE, TRUE, TRUE,TRUE, TRUE, TRUE )
dummydata<- data.frame(Treatment, Estimate, Genus, Sig)
dummydata$Treatment<-factor(dummydata$Treatment, levels = c( "Date_scaled","SeasonDRY", "Maxtemp_12month_scaled", "TB_stageExposed",  "TB_stageDiseased" , "Condition_scaled"))

fig4<-ggplot(final_estimates3, aes(y = Genus, x = Estimate, col = Sig, fill = Sig))+
  facet_wrap(~Treatment, scales = "free_x", nrow = 1, labeller = as_labeller(variable_names))  + 
  geom_errorbarh(aes(xmin = CI_lower, xmax = CI_upper, col = Sig), height = 0, size = 1)+
   geom_point(size = 1.5, col = "black", pch = 21)+
 geom_vline(xintercept = 0, linetype = "longdash")+
  theme_bw(base_size = 12)+
  scale_color_manual(values = c("grey", "dodgerblue3"))+
  scale_fill_manual(values = c("grey", "dodgerblue3"))+
  theme(axis.title.y = element_blank())+
  theme(axis.text.y = element_text(face="bold", size = 8))+
  theme(axis.text.y = element_text(colour=final_estimates3$Colour))+
  theme(legend.position = "none")+
  theme(strip.text.x = element_text(size = 12, face="bold"))+
    geom_blank(data = dummydata)


#########################
#########################
#########################
#########################


pal1<-c(pal[1] , pal[8], pal[5],  pal[7], "black")

classes_to_keep<-c("Bacilli", "Clostridia", "Bacteroidia", "Fusobacteriia")
final_estimates3$Class2<-ifelse(final_estimates3$Class%in% classes_to_keep, final_estimates3$Class, "Other")

unique(final_estimates3$Class)

p1<-ggplot(final_estimates3, aes(y = Genus, x = Estimate, fill = Class2))+
  geom_col()+
  facet_wrap(~Treatment, nrow = 1, scales = "free_x", labeller = as_labeller(variable_names))  + 
  scale_fill_manual(values = pal1)+
  labs(fill = "Class")

legend<-get_legend(p1)

top_panel<-ggarrange( fig4, legend, widths = c(10, 1))+ 
  theme(plot.margin=unit(c(0.2,2.5,0.2,0.5),"cm"))
 


```

### Residual correlation plot

```{r}


cr1<-data.frame(getResidualCor(m_final))#

names(cr1)<-names(ys)
names(cr1)<-abbreviate(names(cr1), minlength = 15)
rownames(cr1)<-abbreviate(rownames(cr1), minlength = 15 )

cr2<-cor_pmat(cr1)

corplot<-ggcorrplot(cr1, 
   hc.order = TRUE,
   outline.col = "white",
   type = "full",
   ggtheme = ggplot2::theme_minimal,
   tl.cex = 7.5,
   p.mat = cr2,
   sig.level = 0.05,
   #show.diag = F,
   insig = "blank",
  # colors = c("#6D9EC1", "white", "#E46726"))
   colors = c("blue", "white", "red"))+
     theme(axis.text.x = element_text(angle = 90, vjust=0.5))+
  theme(plot.margin=unit(c(0.2,0.2,0.2,2),"cm"))



```

### Interaction between rainfall and climate 
```{r fig.width = 15, fig.height = 8.5}

df<-coef(m_climate)
est_df<-data.frame(df$Intercept)
est_df2<-data.frame(df$Xcoef) 
est_df3<-merge(est_df, est_df2, by = 0)


# order genera

row.names(est_df3)<-est_df3$Row.names
est_df3<-est_df3[colnames(ys),]

#put est_df3 into long format

names(est_df3)[1]<- "Genus"
names(est_df3)[2]<- "Intercept"

estimates_df <- gather(est_df3, Treatment, Estimate, names(est_df3)[2]:names(est_df3)[ncol(est_df3)], factor_key=TRUE)

############# extract confindence intervals  ####
############# extract confindence intervals  ####
############# extract confindence intervals  ####
############# extract confindence intervals  ####


confint_df<-data.frame(confint(m_climate))

confint_df<-rbind(confint_df[rownames(confint_df) %like% "Xcoef", ],
confint_df[rownames(confint_df) %like% "Intercept", ])


# add a column with correct variable level
variables<- colnames(est_df3)[3:ncol(est_df3)]
variables<-c(variables, "Intercept")
variables1<-rep(variables, nrow(est_df))
variables2<-variables1[order(match(variables1, variables))]


confint_df$Treatment<-variables2


# column with taxa names. Should be automatically in the correct order but double check

confint_df$Genus<-rep(colnames(ys), length(unique(confint_df$Treatment)))



# now have estimates, confidence intervals, aas seperate data frames, but they are in different formats. Need to massage them into one dataframe for plotting.

merged<-merge(estimates_df, confint_df, by = c("Treatment", "Genus"))


final_estimates<- merged
names(final_estimates)[4]<-"CI_lower"
names(final_estimates)[5]<-"CI_upper"


terms<- c("Climate_catHighTemp_HighRainfall", "Climate_catHighTemp_LowRainfall", 
  "Climate_catLowTemp_LowRainfall", "Intercept" )

final_estimates$Keep<-final_estimates$Treatment %in% terms


final_estimates2<-subset(final_estimates, Keep == T)

intercept<-subset(final_estimates2, Treatment == "Intercept")
intercept$CI_lower <- abs(abs(intercept$CI_lower) - abs(intercept$Estimate))*-1
intercept$CI_upper <- abs(abs(intercept$CI_upper) - abs(intercept$Estimate))
intercept$Estimate<-0

final_estimates2<-rbind(subset(final_estimates2, Treatment != "Intercept"), intercept)

final_estimates2$Treatment<-factor(final_estimates2$Treatment)

## add significance

final_estimates2$Sig<- !data.table::between(0, final_estimates2$CI_lower, final_estimates2$CI_upper)


#########
#########
#########
#########

levels(final_estimates2$Treatment)<- c("LT_HR", "HT_LR", "HT_HR", "LT_LR")
final_estimates2$Treatment<-factor(final_estimates2$Treatment, levels = c("LT_HR","LT_LR",  "HT_HR", "HT_LR"))


interaction_plot<-ggplot(subset(final_estimates2, 
                                Genus == "Bacteroides" |Genus == "Fusobacterium" | 
                                  Genus == "Lactococcus (LAB)"| Genus == "Pediococcus (LAB)" ), 
                         aes(y = Estimate,x = Treatment, col = Treatment, fill = Treatment))+
  facet_wrap(~Genus,  nrow = 1)  + 
  geom_errorbar(aes(ymin = CI_lower, ymax = CI_upper, col = Treatment), width = 0, size = 2)+
   geom_point(size = 4, col = "white", pch = 21)+
 geom_hline(yintercept = 0, linetype = "longdash")+
  theme_bw(base_size = 12)+
    theme(axis.text.x = element_text(angle = 90, vjust=0.5))+
  scale_colour_tableau('Classic Blue-Red 12', labels = c("Low max temp/High rain", "Low max temp/Low rain","High max temp/High rain", "High max temp/Low rain"))+
  scale_fill_tableau('Classic Blue-Red 12', labels = c("Low max temp/High rain", "Low max temp/Low rain","High max temp/High rain", "High max temp/Low rain"))+

  theme(axis.text.y = element_text(face="bold", size = 9))+
  labs(col = "Climate (year prior \nto sampling)", fill = "Climate (year prior \nto sampling)")+
  theme(plot.margin=unit(c(0.2,0.5,0.2,0.3),"cm"))+
  xlab("Climate categories")+
  ylab("Estimate")


lower_panel<- ggarrange( corplot,interaction_plot,  ncol = 2, widths = c(1,1.5), 
                         labels = c("g)", "h)"), align = "h")+
   theme(plot.margin=unit(c(0,0,0.2,0.1),"cm"))


## figure 4
ggarrange(top_panel, lower_panel, ncol = 1, heights = c(1,1))


```

# Fig 5: Network analysis

*Need to do CLR analysis from here as this accounts for bacterial load

```{r chunk32}


#### extract site x species array



meerkat_genus<-microbiome::aggregate_top_taxa(meerkat_filtered, "Genus", top = 50)
clr_scaled <-microbiome::transform(meerkat_genus, transform = "clr")


taxa_to_keep <- c("[Ruminococcus] torques group" , "Arthrobacter", "Bacillus","Bacteroides",
 "Blautia", "Catenisphaera" ,"Christensenellaceae R-7 group", "Clostridium sensu stricto 1",
"Collinsella" ,"Enterococcus" , "Escherichia-Shigella","Fusobacterium",
"Geodermatophilus","Lachnoclostridium", "Lactococcus" ,"Marvinbryantia" ,
"Paeniclostridium","Peptoclostridium" , "Peptococcus" ,"Phascolarctobacterium",
"Romboutsia", "Slackia" ,"Turicibacter" , "Weissella" , 
"Rikenellaceae RC9 gut group", "Helicobacter" , "Parabacteroides", "Lactobacillus",
 "Pediococcus", "Streptococcus" )

meerkat_genus_scaled<-prune_taxa(taxa_to_keep,clr_scaled) # new
meerkat_genus_scaled<-subset_samples(meerkat_genus_scaled, !is.na(Mean_rainfall_12month_NOAA))



phylo_array<-data.frame(t(meerkat_genus_scaled@otu_table@.Data))


#phylo_array[1:5,1:5]
names(phylo_array)<-taxa_names(meerkat_genus_scaled)

phylo_array[1:5,1:5]

########################## 

#### genreate association matrix from the 'real' data
# this uses the package NetCoMi


net_real <- netConstruct(phylo_array, verbose = 0, 
                         measure = "pearson",
                         normMethod = "none",
                         zeroMethod = "none",
                         sparsMethod = "threshold",
                         thresh = 0.3)


props_single <- netAnalyze(net_real, 
                           centrLCC = TRUE,
                           clustMethod = "cluster_fast_greedy",
                           hubPar = "eigenvector",
                           weightDeg = F, 
                           normDeg = F)



plot(props_single, 
     nodeColor = "cluster", 
     layout = "spring",
     repulsion = 1,
     # nodeSize = "eigenvector",
     title1 = F, 
     #showTitle = TRUE,
     cexTitle = 2.3,
     shortenLabels = "simple",
     labelScale = F,
     nodeSize = "fix",
     colorVec = c("skyblue", "orchid",  "pink","orange", "lightgreen"),
     nodeTransp = 0,
     borderCol = "white",
     borderWidth = 3,
     highlightHubs = F ,
     edgeTranspLow = 0)

legend(0.7, 1.1, cex = 1, title = "Direction", 
       legend = c("+","-"), lty = 1, lwd = 1, col = c("#009900","red"), 
       bty = "n", horiz = F)

cluster_memb<-data.frame(props_single$clustering$clust1)

names(cluster_memb)<-"Cluster"
table(cluster_memb$Cluster)

# make singletons their own cluster (not zero)

cluster_memb$Genus<-row.names(cluster_memb)


cluster_memb<-cluster_memb %>% mutate(Cluster2 = case_when(Cluster == 2 ~ "Lactococcus cluster",
                                 Cluster == 1 ~ "Bacteroides cluster",
                                Cluster == 4 ~ "Blautia cluster",
                                 Cluster == 5 ~ "Clostridium cluster",
                                 Cluster == 3 ~ "Bacillus cluster",
                                 Genus == "Escherichia-Shigella" ~ "Escherichia-Shigella",
                                 Genus == "Peptoclostridium" ~ "Peptoclostridium"))

```

### Survival analysis

```{r chunk33}



######### 2) Network approach ############
######### 2) Network approach ############
######### 2) Network approach ############
######### 2) Network approach ############

meerkat_genus<-microbiome::aggregate_top_taxa(meerkat_filtered, "Genus", top = 50)
taxtable<-data.frame(tax_table(meerkat_genus))
taxtable[1:5,1:2]
taxtable_new<-merge(taxtable, cluster_memb, by = "Genus", all.x = T)
taxtable_new<-taxtable_new[,c(1,4)]
taxtable_new$Cluster2<-ifelse(is.na(taxtable_new$Cluster2) ,"None",taxtable_new$Cluster2 )

rownames(taxtable_new)<-taxtable_new$Genus

meerkat_genus@tax_table@.Data[,1]<-taxtable_new$Cluster2


## merge by cluster

cluster_ps<-tax_glom(meerkat_genus, taxrank = "Genus")
tax_table(cluster_ps)


cluster_clr <-microbiome::transform(cluster_ps, transform = "clr")

keep<-taxa_names(cluster_clr)[-1]

cluster_clr<-prune_taxa(keep, cluster_clr)
taxa_names(cluster_clr)
taxa_names(cluster_clr)<- c("Blautia cluster", "Bacillus cluster", "Bacteroides cluster", 
                            "Clostridium cluster", "Lactococcus cluster", "Escherichia","Peptoclostridium" )


cluster_clr<-subset_samples(cluster_clr, !is.na(Mean_rainfall_12month_NOAA))

otutable<-data.frame(t(otu_table(cluster_clr)))

otutable<-data.frame(rescale(otutable, to = c(-1,1)))

### survival data ####
### survival data ####


survival_df<-data.frame(sample_data(cluster_clr))

# calculate if ID is alive 6 months later, TRUE = dead

survival_df$Time_180<-survival_df$SampleDate+180
survival_df$Survival_180<- ifelse(survival_df$Time_180 < survival_df$DeathDate, FALSE,TRUE)


table(survival_df$Survival_180)
survival_df$FollowUp_180<-180

survival_df_network<-merge(survival_df, otutable, by = 0)

#############
#############
#############
#############


survival_df_network$IndividID<-as.character(survival_df_network$IndividID)

survival_df_network$Condition_scaled<-as.numeric(rescale(survival_df_network$Condition_weight, to = c(-1,1)))
survival_df_network$Age_scaled<-as.numeric(rescale(survival_df_network$AgeAtSampling, to = c(-1,1)))
survival_df_network$Rainfall_12month_scaled <-as.numeric(rescale(survival_df_network$Mean_rainfall_12month_NOAA, to = c(-1,1)))
survival_df_network$Maxtemp_12month_scaled <-as.numeric(rescale(survival_df_network$Mean_maxtemp_12month, to = c(-1,1)))


survival_df_network$Season <-case_when(survival_df_network$Season == "WET" ~ "Wet",
                                       survival_df_network$Season == "DRY" ~ "Dry")

survival_df_network<-subset(survival_df_network, !is.na(Survival_180))

###### model #########
###### model #########
###### model #########
###### model #########


fit_network <- coxme::coxme(Surv(FollowUp_180, Survival_180) ~ 

                              Season + 
                              Rainfall_12month_scaled+
                              Maxtemp_12month_scaled+
                              TB_stage + 
                              SocialStatus + 
                              Condition_scaled +
                              Age_scaled+
                              Lactococcus.cluster+
                              Bacteroides.cluster+
                             Blautia.cluster+
                              Bacillus.cluster+
                              Clostridium.cluster+
                             Peptoclostridium+
                              Escherichia+
                              

                             
                              (1 | IndividID),
                            data = survival_df_network)

summary(fit_network)
AIC(fit_network)

### Dredge models 

options(na.action = "na.fail") 

dredge_genus_df<- MuMIn::dredge(fit_network)


write.csv(dredge_genus_df, "dredge_results_clusters.csv")

summary<-summary(model.avg(dredge_genus_df, subset = delta < 4))
summary

#saveRDS(summary, "model_averaging_survival.RDS")
#readRDS("model_averaging_survival.RDS")


### final model 


m_survival <- coxme::coxme(Surv(FollowUp_180, Survival_180) ~ 

                              Season + 
                              Rainfall_12month_scaled+
                              Maxtemp_12month_scaled+
                              TB_stage + 
                              SocialStatus + 
                              Condition_scaled +
                              Age_scaled+
                              Lactococcus.cluster+

                              (1 | IndividID),
                            data = survival_df_network)


labels <- c("TB stage (clinical signs)", "Age", "TB stage (Exposed)", "Social status (Sub)", 
            "Max temp (1 year window)", "Lactococcus cluster", "Season (Wet)", 
            "Rainfall (1 year window)", "Body condition")

labels <- rev(labels)


summary(m_survival)

  
sjPlot::plot_model(m_survival, 
                   axis.lim = c(0.1,20),
                   vline.color = "darkgrey", 
                   sort.est = T,
                   axis.labels = labels,
                   dot.size= 4, 
                   line.size = 1.5, 
                   show.p = T)+
  theme_bw(base_size = 16)+ggtitle("")+ylab("Hazard ratio")



```


